---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crashapi

<!-- badges: start -->
<!-- badges: end -->

The goal of crashapi is to provides functions for downloading data from the National Highway Traffic Safety Administration (NHTSA) [Fatality Analysis Reporting System (FARS) API](https://crashviewer.nhtsa.dot.gov/CrashAPI/). The NHTSA website explains the purpose of the API:

>The NHTSA Crash data Application Programming Interface (API) provides various ways to get crash data collected by DOT NHTSA’s Fatality Analysis Reporting System (FARS) program. The APIs are primarily targeted for developers, programmers or researchers interested in obtaining Crash statistics on Fatal Motor Vehicle Crashes. The API supports multiple data output formats, namely, XML, CSV/XLSX, JSV and JSON. FARS data starting from 2010 onwards is made available through this website. There are different APIs provided to make it easier to query specific data.

Supported APIs include Get Crash List Information (get_fars_crash_list()), Get Crashes By Location (get_fars_crashes()), Get Summary Counts (get_fars_summary()), Get Variables and Get Variable Attributes (`get_fars_vars()`), and Get FARS Data By Year (get_fars_year()).

Multiple formats are only supported for get_fars_year().

The APIs for Get Crash Details, Get Crashes By Vehicle, and Get Crashes By Occupant are not yet supported.

## Installation

You can install the development version of crashapi using the remotes package:

``` r
remotes::install_github("elipousson/crashapi")
```

## Examples

```{r}
library(crashapi)
library(ggplot2)
```

```{r get_fars_crash_list}
# Get crashes in NY from 2019 with 5 to 10 vehicles
get_fars_crash_list(
  start_year = 2019,
  end_year = 2019,
  state = "NY",
  vehicles = c(5,10)
)
```

```{r get_fars_crashes}
# Get crashes for Baltimore County, MD from 2014 to 2015
get_fars_crashes(
  start_year = 2014,
  end_year = 2015,
  state = "MD",
  county = "Baltimore County") |>
  # Show 10 fatal crashes at random
  dplyr::slice_sample(n = 10)
```

```{r map_fars_crashes}
# Get crashes for Baltimore County, MD from 2014 to 2015
# Set geometry to TRUE to return an sf object
crashes <-
  get_fars_crashes(
    start_year = 2014,
    end_year = 2015,
    state = "MD",
    county = "Baltimore County",
    geometry = TRUE
  )

# Map crashes
ggplot() +
  geom_sf(data = dplyr::filter(mapbaltimore::baltimore_msa_counties, namelsad == "Baltimore County"), fill = NA, color = "black") +
  geom_sf(data = crashes, aes(color = TOTALVEHICLES)) +
  theme_void()
```

```{r get_fars_summary}
# Get summary crash count and fatal crash count data for Maryland from 2010 to 2019
get_fars_summary(
  start_year = 2010,
  end_year = 2019,
  state = "MD"
)
```

## Additional Information

The [NHTSA website](https://www-fars.nhtsa.dot.gov/Help/helplinks.aspx) provides additional information on the release data and version status for the FARS Dataset.

| Data Year | File Version | Release Date      |
| --------- | ------------ | ----------------- |
| 2010      | Final        | December 11, 2012 |
| 2011      | Final        | November 13, 2013 |
| 2012      | Final        | December 12, 2013 |
| 2013      | Final        | December 14, 2014 |
| 2014      | Final        | December 18, 2015 |
| 2015      | Final        | December 16, 2016 |
| 2016      | Final        | December 14, 2017 |
| 2017      | Final        | December 18, 2018 |
| 2018      | Final        | June 24, 2021     |
| 2019      | Annual       | June 24, 2021     |

## Related packages and projects

- [stats19](https://github.com/ropensci/stats19) "provides functions for downloading and formatting road crash data" from "the UK’s official road traffic casualty database, STATS19."
- [nzcrash](https://github.com/nacnudus/nzcrash): "An R package to distribute New Zealand crash data in a convenient form."
- [GraphHopper Open Traffic Collection](https://github.com/graphhopper/open-traffic-collection): "Collections of URLs pointing to traffic information portals which contain open data or at least data which is free to use."
